---
title: "Complete Guide to ensembleHTE"
author: "Bruno Fava"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Complete Guide to ensembleHTE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
# Pre-computed fit objects are stored in vignettes/precomputed/.
# Run data-raw/precompute_vignette_fits.R to regenerate them.
PRECOMPUTED <- file.exists("precomputed/fit_hte.rds")
```

# Introduction

## Overview

The `ensembleHTE` package provides a comprehensive framework for learning features of heterogeneous treatment effects (HTEs) in randomized controlled trials (RCTs) using ensemble machine learning methods. The package implements the estimation strategy developed in Fava (2025), which combines predictions from multiple machine learning algorithms using Best Linear Predictor (BLP) weights and repeated K-fold cross-fitting to improve statistical power.

The key innovations over existing methods are **using the entire sample for final inference** rather than holding out data for validation, and **combining multiple machine learning algorithms** to improve predictive performance.

## When to Use This Package

This package is designed for researchers and practitioners who want to:

1. **Detect treatment effect heterogeneity**: Test whether a treatment affects different individuals differently based on their observable characteristics
2. **Characterize beneficiaries**: Identify characteristics of groups of individuals that benefit more (or less) from a treatment
3. **Validate predictions**: Test whether machine learning predictions have genuine out-of-sample predictive power
4. **Predict outcomes**: Use ensemble ML for standard prediction tasks without treatment
5. **Compare targeting strategies**: Evaluate whether unrestricted targeting (rank everyone) outperforms constrained targeting (rank within subgroups)

## Key Concepts

### Heterogeneous Treatment Effects (HTE)

In randomized controlled trials (RCTs), the **Average Treatment Effect (ATE)** measures the mean effect of treatment across the population:

$$\text{ATE} = E[Y_i(1) - Y_i(0)],$$


where $Y_i(1)$ 
and $Y_i(0)$ 
are the potential outcomes for individual $i$ under treatment and control, respectively.

However, treatment effects often vary across individuals. The **Conditional Average Treatment Effect (CATE)** captures this heterogeneity:

$$\text{CATE}(x) = \tau(x) = E[Y_i(1) - Y_i(0) | X_i = x]$$

where $X_i$ represents individual characteristics (covariates).

### Group Average Treatment Effects (GATES)

**GATES** (Chernozhukov et al., 2025) provides a structured way to test for and visualize treatment effect heterogeneity by:

1. Using machine learning to predict individual treatment effects $\hat{\tau}_i$ based on covariates
2. Sorting individuals into quantile groups (e.g., terciles) based on these predictions
3. Estimating the average treatment effect within each group
4. Testing whether high-prediction and low-prediction groups have statistically different effects

### The Ensemble Approach

This package improves upon existing single-algorithm approaches by:

1. **Combining Multiple ML Algorithms**: Uses several algorithms (e.g., random forests, gradient boosting, elastic net) and optimally combines their predictions using BLP weights
2. **K-fold Cross-Fitting**: Trains models on K-1 folds and generates predictions on the held-out fold
3. **Full-Sample Inference**: Uses the entire dataset for final estimation, maximizing statistical power
4. **Repeated Sample-Splitting**: Averages across M repetitions for stability and valid inference. **Important**: For debugging and exploration, use M = 2 or 3 to quickly verify your code works. For final results and publication, use M ≥ 100 at least (use more if you can!). The tests conducted with the methodology and recommended usage require M ≥ 100.

## Installation

Install the development version from GitHub:

```{r installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bfava/ensembleHTE")
```

The package requires R ≥ 4.0.0 and depends on several packages including `grf`, `mlr3`, `data.table`, and `ggplot2`. These will be installed automatically.

For additional ML algorithms, you may want to install optional packages:

```{r optional-packages, eval = FALSE}
install.packages(c("glmnet", "gbm", "ranger"))
```

It is recommended to frequently check for package updates:

```{r check-updates, eval = FALSE}
ensemble_news()
```

# Basic Workflow

## Loading the Data

Throughout this vignette, we use the `microcredit` dataset that ships with
the package. It contains data from a randomized microfinance experiment in
the Philippines where microloans were randomly offered to applicants.
For a gentler, application-oriented introduction to the package using this
same dataset, see the companion article
[Introduction to ensembleHTE](../articles/microcredit-walkthrough.html).
That article walks through a practical analysis step by step, while the
present guide provides comprehensive coverage of every feature and option
available in the package.

```{r load-data}
library(ensembleHTE)
data(microcredit)

dim(microcredit)
head(microcredit[, c("treat", "prop_score", "exp_yrly_end", "hhinc_yrly_end")])

# Variables used throughout this guide
hte_covars <- c("css_creditscorefinal", "lower_window", "own_anybus",
                "max_yearsinbusiness", "css_assetvalue")
profit_covars <- c("css_creditscorefinal", "own_anybus", "css_assetvalue",
                   "age", "gender", "hhinc_yrly_base")
has_loan <- microcredit$treat == 1 & microcredit$loan_size > 0

# Income quintiles (used for restricted comparisons later)
microcredit$hhinc_quintile <- cut(
  microcredit$hhinc_yrly_base,
  breaks = quantile(microcredit$hhinc_yrly_base, probs = seq(0, 1, by = 0.2)),
  labels = paste0("Q", 1:5),
  include.lowest = TRUE
)
```

Key variables:

- `treat`: treatment indicator (1 = loan offered)
- `prop_score`: known propensity score (varies by credit-score window)
- `exp_yrly_end`: yearly business expenses at endline (pesos)
- `hhinc_yrly_end`: yearly household income at endline (pesos)
- `bank_profits_pp`: bank profit per peso lent (only observed for borrowers)

## Fitting an Ensemble HTE Model

The main function for HTE estimation is `ensemble_hte()`:

```{r fit-hte, eval = FALSE}
# Fit ensemble HTE model
# Start with M = 2 for quick debugging, then increase to M >= 100 for final results
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),  # Linear model + causal forest
  M = 2,                         # Use M = 2-3 for debugging, M >= 100 for final results
  K = 3                          # 3-fold cross-fitting
)

# View basic information
print(fit)
```

```{r fit-hte-output, echo = FALSE, eval = PRECOMPUTED}
fit <- readRDS("precomputed/fit_hte.rds")
print(fit)
```

The `print()` method shows:

- Data dimensions and specification
- Algorithms used in the ensemble
- Metalearner strategy (R-learner by default)
- Split-sample parameters (M repetitions, K folds)

For more detailed output:

```{r summary-hte, eval = PRECOMPUTED}
summary(fit)
```

The `summary()` method provides a comprehensive overview including:

- **BLP Analysis**: Estimates of beta1 (ATE) and beta2 (HET) with standard errors and p-values, plus interpretation of whether significant heterogeneity is detected
- **GATES Analysis**: Group average treatment effects showing estimates by quantile group, along with key hypothesis tests (Top-Bottom, Top-All)

## Extracting Individual Treatment Effect Predictions

The `ite()` function extracts the matrix of individualized treatment effect predictions from a fitted HTE model:

```{r ite-extract, eval = PRECOMPUTED}
# Extract ITE matrix: n observations x M repetitions
ite_matrix <- ite(fit)
dim(ite_matrix)
```

Each column corresponds to one repetition of the cross-fitting procedure. For further analysis beyond the package, it is not recommended to average ITE estimates across repetitions, as this would invalidate the inference. Instead, run your analysis separately on each column and then average the results across repetitions.

# Analyzing Treatment Effect Heterogeneity

## Group Average Treatment Effects (GATES)

The `gates()` function computes average treatment effects for groups defined by predicted ITE quantiles:

```{r gates, eval = PRECOMPUTED}
# Compute GATES with 3 groups (terciles)
gates_results <- gates(fit, n_groups = 3)
print(gates_results)
```

### Interpreting GATES Output

The output includes:

**Group Estimates**:

- **Group 1**: Bottom tercile (lowest predicted ITEs)
- **Group 2**: Middle tercile
- **Group 3**: Top tercile (highest predicted ITEs)

Each group shows the estimated average treatment effect for individuals in that group, along with standard errors, t-statistics, and p-values.

**Heterogeneity Tests**:

- **Top-Bottom**: Tests whether the top group has a different average treatment effect than the bottom group. A significant result (p < 0.05) indicates evidence of heterogeneity.
- **Top-All**: Tests whether the top group differs from the overall average. Useful for comparing targeting strategies focusing on high-benefit individuals versus treating everyone.

### Visualizing GATES

```{r plot-gates, eval = PRECOMPUTED, fig.cap="GATES estimates showing treatment effect heterogeneity"}
plot(gates_results)
```

The plot shows:

- Point estimates for each group with 95% confidence intervals
- Blue line indicating the overall ATE
- Groups ordered from lowest to highest predicted benefit

If the confidence intervals don't overlap much (especially between groups 1 and 3), this provides visual evidence of heterogeneity.

## Best Linear Predictor (BLP)

The `blp()` function tests whether the ML predictions capture meaningful heterogeneity:

```{r blp, eval = PRECOMPUTED}
# Compute BLP
blp_results <- blp(fit)
print(blp_results)
```

### Interpreting BLP Output

The BLP regression is:

$$Y_i = \beta_1 (D_i - p_i) + \beta_2 (D_i - p_i)(\hat{\tau}_i - \bar{\hat{\tau}}) + \epsilon_i$$

where $p_i$ is the propensity score (probability of treatment), and $\bar{\hat{\tau}}$ is the average predicted treatment effect.

**Coefficients**:

- **beta1 (ATE)**: Estimates the overall average treatment effect.
- **beta2 (HET)**: Tests heterogeneity. A significant beta2 (p < 0.05) indicates that the ML predictions $\hat{\tau}_i$ capture genuine treatment effect heterogeneity. If beta2 ≈ 1, the predictions are well-calibrated.

You can also add control variables to the BLP regression using the `controls` parameter:

```{r blp-controls, eval = FALSE}
# BLP with additional control variables
blp_with_controls <- blp(fit, controls = c("age", "gender"))
```

### Visualizing BLP

```{r plot-blp, eval = PRECOMPUTED, fig.cap="BLP coefficient estimates"}
plot(blp_results)
```

## Classification Analysis (CLAN)

The `clan()` function characterizes which types of individuals are in the top vs. bottom groups:

```{r clan, eval = PRECOMPUTED}
# Compute CLAN for all covariates
clan_results <- clan(fit, n_groups = 3)
print(clan_results)
```

### Interpreting CLAN Output

CLAN compares covariate means between groups. The print output shows two panels:

**1. Group Means** — Mean (and standard error) of each variable within the Top, Bottom, Else, and All groups.

**2. Differences from Top Group** — Differences in means (Top − Bottom, Top − Else, Top − All) with significance stars and standard errors in parentheses.

This helps answer questions like: "Do individuals predicted to benefit most from treatment have higher X1 on average?"

### Using Analysis Functions with Subsets

All analysis functions (`gates()`, `gavs()`, `clan()`, `gates_restricted()`, `gavs_restricted()`) accept a `subset` argument that restricts which observations are used for the analysis. The `subset` can be a logical vector, an integer index vector, or the string `"train"` (to use the training subset from `train_idx`) or `"all"`. When `subset` is specified, only those observations enter the analysis (e.g., CLAN computes group means only on the subset; GATES runs regressions only on the subset).

Separately, the `group_on` argument controls which observations are used to **define the group cutoffs** (i.e., the quantile thresholds that assign individuals to groups). This is distinct from `subset`: `subset` determines *who is analyzed*, while `group_on` determines *whose predictions define where the group boundaries fall*.

- `group_on = "auto"` (default): Groups are formed using the same population that was used to **train** the ML model. If `ensemble_hte` or `ensemble_pred` are fit without `train_idx`, this means all observations; if they are trained on a subset (via `train_idx`), this means only the training subset. This default ensures that group definitions reflect the population the ML model was designed for.
- `group_on = "all"`: Groups are always formed using **all** observations, regardless of how the ML was trained or which subset is being analyzed.
- `group_on = "analysis"`: Groups are formed using only the observations in the analysis `subset`. This means the group boundaries are re-computed within the subset, so individuals may be assigned to different groups than they would be in the full sample.

```{r clan-subset, eval=FALSE}
# Suppose you want to analyze CLAN only for business owners
subset_biz <- microcredit$own_anybus == 1

# Default (auto): group cutoffs come from the ML training population
clan_all <- clan(fit, n_groups = 3, subset = subset_biz)

# Alternative: form groups only within the business-owner subset
clan_reranked <- clan(fit, n_groups = 3, subset = subset_biz,
                      group_on = "analysis")
```

### Visualizing CLAN

```{r plot-clan, eval = PRECOMPUTED, fig.cap="CLAN results showing characteristics of top vs bottom groups"}
plot(clan_results)
```

The plot shows mean differences with confidence intervals for each variable. Variables with bars that don't include zero are significantly different between groups.

# Prediction Tasks (Without Treatment)

The package also supports standard prediction problems without treatment effects through `ensemble_pred()`. This is useful for:

- Predicting outcomes based on covariates
- Identifying groups with high/low predicted outcomes
- Testing whether ML predictions have genuine predictive power
- Creating predicted scores to use downstream in treatment effect analyses (see [Cross-Outcome Analysis])

## Fitting an Ensemble Prediction Model

```{r fit-pred, eval = FALSE}
# Predict bank profitability using ensemble_pred
fit_pred <- ensemble_pred(
  Y    = "bank_profits_pp",
  X    = profit_covars,
  data = microcredit,
  train_idx  = has_loan,
  algorithms = c("lm", "grf"),
  M = 100, K = 3  # Use M >= 100 for final results
)

print(fit_pred)
summary(fit_pred)
```

```{r fit-pred-output, echo = FALSE, eval = PRECOMPUTED}
fit_pred <- readRDS("precomputed/fit_pred.rds")
print(fit_pred)
summary(fit_pred)
```

## Group Averages (GAVS)

The `gavs()` function computes average outcomes for groups defined by prediction quantiles — the prediction counterpart of GATES:

```{r gavs, eval = PRECOMPUTED}
gavs_results <- gavs(fit_pred, n_groups = 3)
print(gavs_results)
```

This shows whether individuals with high predicted Y actually have higher observed Y on average.

### Visualizing GAVS

```{r plot-gavs, eval = PRECOMPUTED, fig.cap="Group Averages showing predicted vs. observed outcome by group"}
plot(gavs_results)
```

## Best Linear Predictor for Predictions (BLP_PRED)

The `blp_pred()` function tests whether ML predictions have genuine out-of-sample predictive power. This is distinct from `blp()`, which tests for treatment effect heterogeneity.

```{r blp-pred, eval = PRECOMPUTED}
blp_pred_results <- blp_pred(fit_pred)
print(blp_pred_results)
```

### Interpreting BLP_PRED Output

The BLP_PRED regression is:

$$Y_i = \alpha + \beta \hat{Y}_i + \epsilon_i$$

**Coefficients**:

- **Intercept (alpha)**: Should be close to 0 if predictions are well-calibrated.
- **beta**: Tests predictive power. A significant beta (p < 0.05) indicates that the ML predictions $\hat{Y}_i$ carry genuine predictive content. If beta ≈ 1, the predictions are well-calibrated in magnitude.

### Visualizing BLP_PRED

```{r plot-blp-pred, eval = PRECOMPUTED, fig.cap="BLP_PRED coefficient estimates"}
plot(blp_pred_results)
```

# Customizing the Ensemble

This section covers the options available for customizing how the ensemble is fitted. These options apply to both `ensemble_hte()` and `ensemble_pred()` unless stated otherwise.

## Metalearner Strategies

The `metalearner` parameter in `ensemble_hte()` controls the strategy for estimating individual treatment effects. The package supports four strategies:

### R-learner (Default)

The R-learner (Nie & Wager, 2021) uses residual-on-residual regression:

1. Estimate the conditional mean $\hat{m}(x) = E[Y|X=x]$ using the specified base algorithm
2. Compute residuals: $\tilde{Y}_i = Y_i - \hat{m}(X_i)$ and $\tilde{D}_i = D_i - p_i$
3. Estimate the CATE $\hat{\tau}(x)$ by regressing $\tilde{Y}$ on $\tilde{D}$ (via a causal forest or weighted regression)

**Note on propensity scores**: The original R-learner framework (Robinson, 1988; Nie & Wager, 2021) estimates the propensity score $e(x) = E[D|X]$ as a nuisance parameter. Since **this package is designed for RCTs**, where the propensity score is known by design, the implementation uses the **known propensity score** $p_i$ directly rather than estimating it. This avoids an unnecessary estimation step and uses the exact treatment assignment probabilities from the experimental protocol.

The `r_learner` parameter controls the algorithm used for the CATE estimation step. By default, it uses `"grf"` (causal forest), but any mlr3 learner can be specified:

```{r metalearner-r, eval = FALSE}
fit_r <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  metalearner = "r",
  r_learner = "grf",  # Use causal forest for CATE estimation (default)
  M = 100, K = 3  # Use M >= 100 for final results
)
```

### T-learner

Trains separate models for treated and control groups:

$$\hat{\tau}(x) = \hat{\mu}_1(x) - \hat{\mu}_0(x)$$

```{r metalearner-t, eval = FALSE}
fit_t <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  metalearner = "t",
  M = 100, K = 3
)
```

### S-learner

Trains a single model with treatment as a feature:

$$\hat{\mu}(x, d) = E[Y | X=x, D=d]$$
$$\hat{\tau}(x) = \hat{\mu}(x, 1) - \hat{\mu}(x, 0)$$

```{r metalearner-s, eval = FALSE}
fit_s <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  metalearner = "s",
  M = 100, K = 3
)
```

### X-learner

A two-stage approach (Künzel et al., 2019) that imputes counterfactual outcomes:

```{r metalearner-x, eval = FALSE}
fit_x <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  metalearner = "x",
  M = 100, K = 3
)
```

**Recommendation**: The R-learner with `r_learner = "grf"` is generally a good default. T-learner works well with sufficient sample sizes in each arm. S-learner can underestimate heterogeneity if the base learner struggles to detect treatment-covariate interactions.

## Available Algorithms

The `algorithms` parameter accepts algorithms from two sources:

### grf Package

Use `"grf"` for generalized random forest (Athey, Tibshirani & Wager, 2019):

```{r algo-grf, eval = FALSE}
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("grf"),
  M = 3, K = 3
)
```

### mlr3 Learners

Any regression learner supported by mlr3 can be used. Specify just the algorithm name.
See the [mlr3learners documentation](https://mlr3learners.mlr-org.com) for
recommended learners, or the [full list of available learners](https://mlr-org.com/learners.html)
for all options. 
Some examples:

| Algorithm | Name | Description |
|-----------|------|-------------|
| Linear Model | `"lm"` | Standard linear regression |
| Random Forest | `"ranger"` | Fast random forest implementation |
| Elastic Net | `"glmnet"` | L1/L2 regularized regression |
| Gradient Boosting | `"xgboost"` | XGBoost gradient boosting |
| Neural Network | `"nnet"` | Single-layer neural network |
| K-Nearest Neighbors | `"kknn"` | K-nearest neighbors |
| Support Vector Machine | `"svm"` | Support vector machine |

### Ensemble Recommendations

For most applications, we recommend using **3-4 algorithms** (at most 5). A good general-purpose ensemble:

```{r algo-recommended, eval = FALSE}
# Good general-purpose ensemble (4 algorithms)
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("glmnet", "grf", "nnet", "xgboost"),
  M = 3, K = 3  # Use M >= 100 for final results
)
```

This combines:

- **glmnet**: Elastic Net
- **grf**: Generalized Random Forest
- **nnet**: Neural Network
- **xgboost**: Gradient Boosting

## Hyperparameter Tuning

Hyperparameter tuning applies only to **mlr3 learners** (e.g., `"ranger"`, `"glmnet"`, `"xgboost"`). The `"grf"` algorithm uses its own built-in tuning and is not affected by the `tune` parameter.

Enable automatic tuning by setting `tune = TRUE`. The function has sensible defaults, so no additional parameters are required:

```{r tuning-simple, eval = FALSE}
# Simple usage with defaults
fit_tuned <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("ranger", "glmnet"),
  tune = TRUE,  # Just set this to TRUE - defaults handle the rest
  M = 3, K = 3
)
```

### Simple Tuning Mode

You can customize the built-in tuning process by passing a list of scalar parameters via `tune_params`:

```{r tuning-custom, eval = FALSE}
# Customized tuning parameters (optional)
fit_tuned <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("ranger", "glmnet"),
  tune = TRUE,
  tune_params = list(
    time = 60,              # Max tuning time in seconds per algorithm (default: 30)
    cv_folds = 3,           # Number of CV folds for tuning (default: 3)
    stagnation_iters = 250, # Stop if no improvement for this many iterations (default: 250)
    stagnation_threshold = 0.01  # Minimum improvement threshold (default: 0.01)
  ),
  M = 3, K = 3
)
```

### Advanced Tuning Mode

For full control over the tuning process, you can pass mlr3tuning objects directly. This lets you specify a custom tuner, terminator, resampling strategy, search space, and measure:

```{r tuning-advanced, eval = FALSE}
# Advanced: pass mlr3tuning objects for full control
fit_tuned <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("ranger", "glmnet"),
  tune = TRUE,
  tune_params = list(
    tuner       = mlr3tuning::tnr("grid_search", resolution = 5),
    terminator  = mlr3tuning::trm("evals", n_evals = 50),
    resampling  = mlr3::rsmp("holdout"),
    measure     = "regr.rsq"
  ),
  M = 3, K = 3
)
```

See the [mlr3tuning documentation](https://mlr3tuning.mlr-org.com) for details on available tuners, terminators, and search spaces.

**Note**: Tuning significantly increases computation time. For exploratory analysis, use `tune = FALSE` and only enable tuning for final results.

## Custom Learner Parameters

You can pass algorithm-specific hyperparameters using the `learner_params` argument. This applies only to **mlr3 learners** and is ignored for `"grf"` (which uses its own internal defaults). The argument is a named list where each element is named after an algorithm and contains a list of parameter-value pairs. The parameter names follow the **mlr3 parameter naming convention** — to see which parameters are available for a given learner, run `mlr3::lrn("regr.<algorithm>")$param_set`:

```{r learner-params, eval = FALSE}
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("ranger", "xgboost"),
  learner_params = list(
    ranger  = list(num.trees = 500, min.node.size = 5),
    xgboost = list(nrounds = 200, max_depth = 4)
  ),
  M = 3, K = 3
)
```

These values override algorithm-specific defaults without invoking the tuning machinery. Useful when you already know good hyperparameter values for your data.

## Ensemble Strategy

By default, the package combines algorithm predictions using BLP weights estimated via cross-validation (`ensemble_strategy = "cv"`). You can switch to simple averaging:

```{r ensemble-strategy, eval = FALSE}
# Simple average of all algorithm predictions (equal weights)
fit_avg <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf", "ranger"),
  ensemble_strategy = "average",  # Equal-weighted average instead of BLP
  M = 3, K = 3
)
```

The cross-validation BLP strategy (`"cv"`, default) is generally preferred, as it learns data-adaptive weights that upweight better-performing algorithms.

## Propensity Scores

This package is designed for **randomized controlled trials (RCTs)**, where the propensity score (probability of treatment) is known by design.

**Default behavior**: If you don't specify `prop_score`, the function assumes constant propensity equal to the sample treatment proportion. This is appropriate for simple randomization with equal assignment probabilities.

```{r propensity-default, eval = FALSE}
# For standard 50/50 randomization, no need to specify prop_score
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data = microcredit,
  M = 3, K = 3
)  # Default: prop_score = mean(D) for all observations
```

**When to specify `prop_score`**: You only need to provide `prop_score` when the assignment probability **varies across individuals**. If everyone has the same probability (even if not 50%), the default will correctly use the sample treatment fraction:

```{r propensity-known, eval = FALSE}
# In the microcredit experiment, propensity scores vary by credit-score window.
# Pass the column name to use them:
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",  # Required when probabilities vary
  M = 3, K = 3
)
```

**Important**: Since this package is for RCTs, you should always **know** the propensity scores from your experimental design. Do not estimate propensity scores — use the true assignment probabilities from your randomization protocol.

## Scaling and Data Preprocessing

By default, covariates are standardized before being passed to ML algorithms (`scale_covariates = TRUE`). Two details are worth noting:

- **Binary variables** (columns with only 0 and 1 values) are **never scaled**, since scaling would remove their natural interpretation
- **Scaling is fit on training data only**: When using `train_idx`, scaling parameters (mean and standard deviation) are computed from the training observations and applied to all observations, preventing data leakage
- **Scaling is only used for ML training and prediction**: The standardized covariates are used internally to train models and generate predictions. All downstream analysis functions (`gates()`, `blp()`, `clan()`, `gavs()`, etc.) operate on the **original, unscaled data**. For example, CLAN reports group means in the original units of each variable.

Factor and character columns are automatically handled: they are converted to factors and, for algorithms that require numeric input (such as `grf`), one-hot encoded using dummy variables.

You can disable scaling if your covariates are already on comparable scales:

```{r scale-covariates, eval = FALSE}
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  scale_covariates = FALSE,  # Disable automatic scaling
  M = 3, K = 3
)
```

## Parallel Processing

Speed up computation by parallelizing across repetitions:

```{r parallel, eval = FALSE}
# Use 4 cores for faster computation
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),
  M = 100,
  K = 3,
  n_cores = 4  # Parallelize across 4 cores
)
```

## Combining Fits from Multiple Sessions

You can combine fits from separate runs using `combine_ensembles()`. This is useful in two common scenarios: (1) you ran an initial analysis and later want to **add more repetitions** to increase M without re-running everything, or (2) you split work across multiple machines or cluster jobs for faster computation:

```{r combine-fits, eval = FALSE}
# Session 1: Run on server 1
fit1 <- ensemble_hte(
  Y = "exp_yrly_end", D = "treat", X = hte_covars,
  data = microcredit, prop_score = "prop_score",
  M = 50, K = 3
)
saveRDS(fit1, "fit_server1.rds")

# Session 2: Run on server 2
fit2 <- ensemble_hte(
  Y = "exp_yrly_end", D = "treat", X = hte_covars,
  data = microcredit, prop_score = "prop_score",
  M = 50, K = 3
)
saveRDS(fit2, "fit_server2.rds")

# Later: Combine results
fit1 <- readRDS("fit_server1.rds")
fit2 <- readRDS("fit_server2.rds")
fit_combined <- combine_ensembles(fit1, fit2)
print(fit_combined)  # Shows M = 100
```

For large-scale cluster computing with many nodes (e.g., 100 jobs), you can combine multiple fits using `Reduce()`:

```{r combine-many-fits, eval = FALSE}
# Example: combining fits from 100 cluster jobs
# Each job saved as fit_1.rds, fit_2.rds, ..., fit_100.rds

# Load all fit files
fit_files <- list.files(pattern = "fit_[0-9]+\\.rds$")
fits <- lapply(fit_files, readRDS)

# Combine all fits at once
fit_final <- Reduce(combine_ensembles, fits)
print(fit_final)  # Shows total M across all jobs
```

# Cross-Outcome Analysis

One of the most powerful features of this package is the ability to **analyze outcomes different from the one used for prediction**. This enables several important research designs that go beyond standard HTE analysis.

## Overview: Decoupling Prediction from Analysis

The key insight is that you can:

1. **Train predictions on one variable** (using `ensemble_hte()` or `ensemble_pred()`)
2. **Form groups** based on those predictions
3. **Analyze treatment effects or averages** on a completely different outcome

This decoupling opens up many possibilities:

| Prediction Target | Analysis Target | Use Case |
|-------------------|-----------------|----------|
| ITE on outcome Y | Treatment effect on Z | Do predicted beneficiaries on Y also benefit on Z? |
| Predicted Y (no treatment) | Treatment effect on Z | Do individuals with high predicted Y experience larger treatment effects on Z? |
| ITE on outcome Y | Average of endline Z by group | What are the post-treatment characteristics of predicted beneficiaries? |

**Note on the third case**: This is distinct from CLAN, which examines baseline covariates X. Here, Z is an **endline outcome** measured after treatment — something that might itself be affected by treatment. For example, you might want to know: "Among those predicted to benefit most on test scores (Y), what is their average college enrollment rate (Z)?" This helps characterize beneficiaries using post-treatment outcomes.

## Example 1: Predict ITE, Analyze Treatment Effects on Different Outcome

Suppose you estimate treatment effects on business expenses, but also want to
know if the same borrowers who see larger effects on expenses also experience
larger effects on household income.

```{r cross-outcome-ite-gates, eval = FALSE}
# Step 1: Fit HTE model predicting treatment effects on business expenses
fit_exp <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Step 2: Analyze GATES on the DIFFERENT outcome (household income)
# Groups are formed by predicted ITE on business expenses
# But treatment effects are estimated on household income
gates_hhinc <- gates(fit_exp, outcome = "hhinc_yrly_end")
print(gates_hhinc)
```

```{r cross-outcome-ite-gates-output, echo = FALSE, eval = PRECOMPUTED}
# fit_exp is equivalent to fit (same Y, D, X) — reuse the loaded fit
gates_hhinc <- gates(fit, outcome = "hhinc_yrly_end")
print(gates_hhinc)
```

This answers: **"Do individuals with high predicted treatment effects on business expenses also experience high treatment effects on household income?"**

If the top-bottom difference is significant, it suggests that the heterogeneity you detected on expenses "transfers" to income — the same individuals are affected on both outcomes.

## Example 2: Predict Outcome (No Treatment), Analyze Treatment Effects

Sometimes you want to use a **prediction** (without treatment structure) to see
if it predicts who benefits from treatment on a different outcome. In the
microcredit setting, we can predict bank profitability and then ask whether
borrowers the bank would rank as most profitable also see larger treatment
effects on household income.

```{r cross-outcome-pred-gates, eval = FALSE}
# Step 1: Predict bank profits using ensemble_pred (no treatment structure)
fit_pred <- ensemble_pred(
  Y    = "bank_profits_pp",
  X    = profit_covars,
  data = microcredit,
  train_idx  = has_loan,
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Step 2: Use gates() to analyze treatment effects on household income
# Groups are formed by predicted bank profits
# Treatment effects are estimated on household income
gates_from_pred <- gates(
  fit_pred,
  outcome    = "hhinc_yrly_end",
  treatment  = "treat",
  prop_score = "prop_score"
)
print(gates_from_pred)
```

```{r cross-outcome-pred-gates-output, echo = FALSE, eval = PRECOMPUTED}
# fit_pred is already loaded — reuse it
gates_from_pred <- gates(
  fit_pred,
  outcome    = "hhinc_yrly_end",
  treatment  = "treat",
  prop_score = "prop_score"
)
print(gates_from_pred)
```

This answers: **"Do individuals with high predicted bank profits have different treatment effects on household income?"**

## Example 3: Predict ITE, Analyze Group Averages of Another Variable

You can also examine the **average level** (not treatment effect) of a different variable across groups defined by predicted treatment effects. This extends the CLAN analysis to any variable.

```{r cross-outcome-ite-gavs, eval = FALSE}
# Step 1: Fit HTE model on business expenses
fit_exp <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Step 2: Look at average bank profits by ITE group using gavs()
# Groups are formed by predicted ITE on business expenses
# We see how bank profits vary across those groups
gavs_profits <- gavs(fit_exp, outcome = "bank_profits_pp", subset = has_loan)
print(gavs_profits)
plot(gavs_profits)
```

```{r cross-outcome-ite-gavs-output, echo = FALSE, eval = PRECOMPUTED, fig.alt = "Bar chart of average bank profits by ITE group"}
gavs_profits <- gavs(fit, outcome = "bank_profits_pp", subset = has_loan)
print(gavs_profits)
plot(gavs_profits)
```

This answers: **"What are the average bank profits among borrowers with high vs. low predicted treatment effects on business expenses?"**

This is useful for:

- **Characterizing beneficiaries with endline outcomes**: Beyond CLAN (which looks at baseline covariates X), you can examine post-treatment outcomes like employment status, health indicators, or other endline measures
- **Understanding mechanisms**: If predicted beneficiaries on Y also have higher average Z, this might suggest a pathway through which treatment effects operate

# Training on Subsets

A powerful feature of both `ensemble_pred()` and `ensemble_hte()` is the `train_idx` parameter, which allows you to **train on a subset of observations while generating predictions for everyone**. This is essential when an outcome is only observed for some units.

## Motivation: Outcomes Only Observed for Treated

In many experiments, certain outcomes are **only observed for treated individuals**. For example:

- **Microloan program**: Lender profits (interest payments, repayment rates) are only observed for borrowers who received loans
- **Educational intervention**: Advanced test scores only measured for students who completed the program

However, you may want to:

1. Predict that outcome for everyone (including controls)
2. Use those predictions to analyze treatment effects on a different outcome that IS observed for everyone

## Example: Predict Treated-Only Outcome, Analyze Universal Outcome

In the microcredit experiment:

- **Bank profits** (`bank_profits_pp`) are only observed for borrowers who
  received a loan and took it up — the bank only earns revenue on disbursed
  loans.
- **Household income** (`hhinc_yrly_end`) is observed for everyone (treated
  and control).

We want to know: "Do individuals with high predicted bank profits have
different treatment effects on household income?"

```{r subset-training, eval = FALSE}
# Step 1: Train prediction model on loan recipients only (where profits observed)
# But generate predicted bank profits for EVERYONE
fit_profits <- ensemble_pred(
  Y    = "bank_profits_pp",
  X    = profit_covars,
  data = microcredit,
  train_idx  = has_loan,  # Only train on loan recipients
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Predictions are now available for ALL 1,113 applicants (including controls)
summary(fit_profits)
```

```{r subset-training-output, echo = FALSE, eval = PRECOMPUTED}
# fit_profits is equivalent to fit_pred (same call) — reuse loaded object
fit_profits <- fit_pred
summary(fit_profits)
```

## Using Subset Predictions for Treatment Effect Analysis

Now you can use the predicted bank profits (which are available for everyone)
to analyze treatment effects on household income:

```{r subset-gates, eval = PRECOMPUTED}
# Step 2: Analyze treatment effects on income, grouped by predicted bank profits
# This answers: "Do individuals with high predicted bank profits have different
# treatment effects on household income?"
gates_by_profits <- gates(
  fit_profits,
  outcome    = "hhinc_yrly_end",
  treatment  = "treat",
  prop_score = "prop_score"
)
print(gates_by_profits)
```

This design is valuable because:

1. **Bank profits were only observed for loan recipients**
2. By **predicting bank profits for everyone**, you can now examine whether
   predicted-profit groups have different treatment effects on income
3. This enables **targeting based on a predicted characteristic** (bank profits)
   that was only measurable in treated individuals

## Summary: Workflow for Subset Training

1. Identify the outcome that is only observed for a subset (e.g., treated only)
2. Use `ensemble_pred()` with `train_idx` to train on that subset but predict for all
3. Use `gates()` or `gavs()` with the `outcome` parameter to analyze a different variable that IS observed for everyone
4. Optionally, use `treatment` parameter to specify the treatment variable

```{r subset-workflow-summary, eval = FALSE}
# Complete workflow:
fit_subset <- ensemble_pred(
  Y    = "bank_profits_pp",
  X    = profit_covars,
  data = microcredit,
  train_idx  = has_loan,  # Train only on loan recipients
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Option A: GATES - treatment effects on income by predicted profit groups
gates_results <- gates(
  fit_subset,
  outcome    = "hhinc_yrly_end",
  treatment  = "treat",
  prop_score = "prop_score"
)

# Option B: GAVS - average income by predicted profit groups (no treatment structure)
gavs_results <- gavs(fit_subset, outcome = "hhinc_yrly_end")
```

## HTE with Subset Training (Multi-Arm Trials)

The `train_idx` parameter is also available in `ensemble_hte()`, which is
useful for **multi-arm trials** or when you want to estimate treatment effects
on a subset but predict for everyone. For example, in a three-arm trial you
could estimate treatment effects between two arms (A vs. B), then predict ITEs
for all individuals including those in arm C:

```{r hte-train-idx, eval = FALSE}
# Simulate a three-arm trial: Control (C), Treatment A, Treatment B
set.seed(42)
N <- 600
arm <- sample(c("C", "A", "B"), N, replace = TRUE)
X1 <- rnorm(N)
X2 <- rnorm(N)
# Treatment A has heterogeneous effects driven by X1
Y <- 1 + 0.5 * X2 + (arm == "A") * (2 + 1.5 * X1) +
     (arm == "B") * 1 + rnorm(N)

three_arm <- data.frame(
  Y = Y, arm = arm, X1 = X1, X2 = X2,
  D_A = as.integer(arm == "A")  # Treatment indicator: A vs C
)

# Estimate HTE for A vs. C, predict for everyone (including arm B)
fit_A <- ensemble_hte(
  Y = "Y",
  D = "D_A",
  X = c("X1", "X2"),
  data      = three_arm,
  train_idx = three_arm$arm %in% c("A", "C"),  # Train on arms A and C only
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# GATES on training observations (arms A and C)
gates(fit_A)

# GATES on all observations (including arm B)
gates(fit_A, subset = "all")
```

```{r hte-train-idx-output, echo = FALSE, eval = PRECOMPUTED}
fit_subset_hte <- readRDS("precomputed/fit_subset_hte.rds")
gates(fit_subset_hte)
gates(fit_subset_hte, subset = "all")
```

### Comparing Heterogeneity Across Arms

A natural follow-up in multi-arm trials is: **"Do individuals predicted to benefit most from treatment A also benefit most from treatment B?"** You can answer this by fitting separate HTE models for each arm comparison and then using cross-outcome analysis:

```{r multi-arm-cross, eval = FALSE}
# Step 1: Also estimate HTE for B vs. C
three_arm$D_B <- as.integer(three_arm$arm == "B")

fit_B <- ensemble_hte(
  Y = "Y",
  D = "D_B",
  X = c("X1", "X2"),
  data      = three_arm,
  train_idx = three_arm$arm %in% c("B", "C"),  # Train on arms B and C only
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Step 2: Use fit_A to group individuals by predicted benefit from A,
# then analyze treatment effects of B (on the B-vs-C comparison sample)
gates_B_by_A <- gates(
  fit_A,
  outcome   = "Y",
  treatment = "D_B",
  subset    = three_arm$arm %in% c("B", "C")
)
print(gates_B_by_A)

# Step 3: Conversely, group by predicted benefit from B,
# then analyze treatment effects of A
gates_A_by_B <- gates(
  fit_B,
  outcome   = "Y",
  treatment = "D_A",
  subset    = three_arm$arm %in% c("A", "C")
)
print(gates_A_by_B)
```

If the top-bottom difference in `gates_B_by_A` is significant, it means that individuals predicted to benefit most from treatment A **also** benefit more from treatment B — suggesting a common dimension of heterogeneity across treatments. If it is not significant, the two treatments may affect different types of individuals.

# Comparing Targeting Strategies

A key application of HTE estimation is developing **targeting rules**: who should receive treatment? The package provides tools to compare different targeting strategies.

## Unrestricted vs. Restricted GATES

Consider a policy question: Should we target individuals globally based on predicted benefit, or should we target within subgroups (e.g., ensuring each income group receives its "fair share" of treatment)?

The `gates_restricted()` function formally tests whether these strategies yield different outcomes. Using the model fitted above and income quintiles (created in the data-loading step):

```{r gates-compare, eval = PRECOMPUTED}
# Compare strategies
comparison <- gates_restricted(fit, restrict_by = "hhinc_quintile", n_groups = 3)
print(comparison)
```

### Interpreting Comparison Results

The output shows three sets of results:

1. **Unrestricted GATES**: Groups formed by ranking predictions across the full sample
2. **Restricted GATES**: Groups formed by ranking predictions within each group defined by `restrict_by` (e.g., income quintiles)
3. **Difference**: The difference between strategies with properly computed standard errors

Key tests:

- **Top-Bottom Diff**: Tests whether the "top-bottom gap" is larger with unrestricted targeting. A significant positive difference suggests unrestricted targeting better identifies high-benefit individuals.
- **Group-level Differences**: Shows whether each group's average differs between strategies.

### Visualizing Strategy Comparison

```{r plot-gates-compare, eval = PRECOMPUTED, fig.cap="Comparison of unrestricted vs. restricted targeting strategies"}
plot(comparison)
```

The plot shows both strategies side-by-side, making it easy to see whether they produce meaningfully different results.

## Unrestricted vs. Restricted GAVS

The same comparison framework applies to prediction tasks via `gavs_restricted()`:

```{r gavs-compare, eval = PRECOMPUTED, fig.alt = "Comparison of unrestricted vs restricted prediction strategies"}
# Compare prediction strategies by income quintile
gavs_comparison <- gavs_restricted(
  fit_pred,
  restrict_by = "hhinc_quintile",
  n_groups = 3,
  subset = "train"
)
print(gavs_comparison)
plot(gavs_comparison)
```

## When to Use Restricted Targeting

Restricted targeting may be preferred when:

1. **Equity concerns**: Ensuring each subgroup receives some treatment
2. **Fairness requirements**: Regulatory or ethical requirements for balanced allocation

The comparison tests help quantify the **efficiency cost** of restricted targeting: how much predicted benefit is lost by not targeting globally?

# Panel Data Support

## Overview

The `ensembleHTE` package supports panel data (repeated observations per individual) through the `individual_id` argument. When panel data is detected, two adjustments are made automatically:
  
1. **Cluster-aware fold splitting**: During K-fold cross-fitting, all observations from the same individual are assigned to the same fold. This prevents data leakage across folds (training on one time period for an individual while predicting another).
2. **Cluster-robust standard errors**: All statistical tests (GATES, BLP, GAVS, CLAN, and restricted comparisons) use clustered standard errors at the individual level, properly accounting for within-individual correlation.

## Example: Panel Data with ensemble_hte

The `microcredit` dataset is cross-sectional, so this example uses simulated
panel data. Suppose 200 individuals are each observed over 4 time periods,
with treatment assigned at the individual level:

```{r panel-hte, eval = FALSE}
# Simulate panel data: 200 individuals x 4 periods
set.seed(123)
N <- 200; T_periods <- 4
panel_data <- data.frame(
  id      = rep(1:N, each = T_periods),
  period  = rep(1:T_periods, times = N),
  D       = rep(rbinom(N, 1, 0.5), each = T_periods),
  X1      = rnorm(N * T_periods),
  X2      = rnorm(N * T_periods)
)
panel_data$Y <- 2 + panel_data$D * (1 + panel_data$X1) +
  0.5 * panel_data$X2 + rnorm(N * T_periods)

# Fit ensemble HTE model with panel data support
fit <- ensemble_hte(
  Y = "Y",
  D = "D",
  X = c("X1", "X2"),
  data = panel_data,
  individual_id = "id",  # Specify the individual identifier
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# All analysis functions automatically use clustered SEs
gates_results <- gates(fit, n_groups = 3)
blp_results <- blp(fit)
clan_results <- clan(fit)
```

## Example: Panel Data with ensemble_pred

```{r panel-pred, eval = FALSE}
# Fit ensemble prediction model with panel data
fit_pred <- ensemble_pred(
  Y = "Y",
  X = c("X1", "X2"),
  data = panel_data,
  individual_id = "id",
  algorithms = c("lm", "grf"),
  M = 3, K = 3
)

# Analysis functions automatically use clustered SEs
gavs_results <- gavs(fit_pred, n_groups = 3)
blp_pred_results <- blp_pred(fit_pred)
```

## How It Works

When `individual_id` is provided:

- **Fold assignment**: The K-fold split is performed at the individual level. If there are $N$ individuals, each fold contains approximately $N/K$ individuals (with all their time periods). This ensures that the same individual never appears in both training and test sets within a fold.
- **Standard errors**: All regressions use `sandwich::vcovCL()` with clustering at the individual level instead of `sandwich::vcovHC()`. CLAN uses a regression-based approach for cluster-robust SEs when clustering is active. Cross-regression covariance calculations (used in `gates_restricted()` and `gavs_restricted()`) also account for clustering.

**Important**: The `individual_id` can be specified as a column name (quoted or unquoted) or as a vector. The column does not need to be included among the covariates in the formula.

# Best Practices and Recommendations

## Sample Size Considerations

The ensemble approach works best with sufficient sample size. We recommend using **3-4 algorithms** (at most 5) and **K = 3** folds for most applications:

| n | Algorithms | K | Groups | Notes |
|---|------------|---|--------|-------|
| < 200 | 1-2 | 2 | 3 | Results may be unstable; interpret with caution |
| 200-500 | 1-2 | 3 | 3 | K = 3 recommended; K = 5 may be unstable |
| 500-2000 | 3-4 | 3-5 | 3 | Standard usage |
| > 2000 | 3-5 | 3-5 | 3-5 | Can use K ≥ 5 for large samples |

## Choosing Parameters

### Number of Repetitions (M)

- **Debugging**: M = 2-3 for quick code verification
- **Final analysis**: **M ≥ 100** for reproduciblility and valid inference

**Important**: The empirical performance of this package was tested using M ≥ 100. Results with smaller M are useful only for debugging and should not be used for publication.

### Number of Folds (K)

- **K = 3**: Recommended for most applications, including small to moderate samples
- **K = 5**: Can use for large samples (n > 2000), but typically unnecessary
- **K = 10**: Not recommended — increases computation with minimal benefit

### Number of Groups

- **n_groups = 3**: Standard choice (terciles); recommended for most analyses
- **n_groups = 5**: More granular, but requires larger samples for stable estimates
- More groups = more precise targeting but higher variance in estimates

## Computational Performance Tips

Total computation time scales roughly as `M × K × length(algorithms)`. 

**Tips for faster runs:**

1. **Parallelization**: Set `n_cores > 1` to distribute repetitions across CPU cores. This provides near-linear speedup.
2. **Start small**: Use `M = 2`, `K = 3`, and 2 algorithms to verify your code works before scaling up.
3. **Tuning**: `tune = TRUE` can increase per-algorithm time significantly. Use only for final results.
4. **Cluster computing**: Use `combine_ensembles()` to split work across machines and combine later.
5. **Algorithm choice**: `lm` (OLS) is fast, but less powerful. `grf` (regression forest) is often faster than `xgboost` (especially with tuning), and typically faster than `nnet` and `svm`. 

## Interpreting Results

### What Indicates Strong Heterogeneity?

1. **Significant Top-Bottom difference** in GATES (p < 0.05)
2. **Significant beta2** in BLP (p < 0.05)
3. **GATES plot** showing non-overlapping confidence intervals between top and bottom groups

### What If No Heterogeneity Is Found?

No significant heterogeneity could mean:

1. **True homogeneity**: Treatment affects everyone similarly
2. **Underpowered**: Sample too small to detect heterogeneity
3. **Wrong covariates**: Heterogeneity exists but along unmeasured dimensions
4. **Model limitations**: ML algorithms unable to capture complex patterns

Consider: excluding noisy/irrelevant covariates, adding relevant covariates, or trying different algorithms. But be careful with specification searching to avoid false positives!

# Conclusion

The `ensembleHTE` package provides a comprehensive toolkit for learning features of heterogeneous treatment effects. By combining multiple ML algorithms with repeated cross-fitting, it achieves higher statistical power.

Key takeaways:

1. **Use `ensemble_hte()` for treatment effect heterogeneity** in RCTs
2. **Use `ensemble_pred()` for prediction problems** without treatment structure
3. **Use `gates()` and `blp()` to test for heterogeneity**; use `gavs()` and `blp_pred()` for prediction validation
4. **Use `clan()` to characterize high/low benefit groups**
5. **Use `gates_restricted()` and `gavs_restricted()` to evaluate targeting strategies**
6. **Use cross-outcome analysis** to study how predictions on one variable relate to effects on another
7. **Use `train_idx`** to train on subsets while predicting for everyone

For questions, bug reports, or feature requests, please visit the [GitHub repository](https://github.com/bfava/ensembleHTE).

# References

Chernozhukov, V., Demirer, M., Duflo, E., & Fernández-Val, I. (2025). Fisher-Schultz Lecture: Generic Machine Learning Inference on Heterogeneous Treatment Effects in Randomized Experiments, with an Application to Immunization in India. *Econometrica*, 93(4), 1121-1164.

Fava, B. (2025). Training and Testing with Multiple Splits: A Central Limit Theorem for Split-Sample Estimators. *arXiv preprint arXiv:2511.04957*.

Künzel, S.R., Sekhon, J.S., Bickel, P.J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *Proceedings of the National Academy of Sciences*, 116(10), 4156-4165.

Nie, X., & Wager, S. (2021). Quasi-Oracle Estimation of Heterogeneous Treatment Effects. *Biometrika*, 108(2), 299-319.

Robinson, P.M. (1988). Root-N-Consistent Semiparametric Regression. *Econometrica*, 56(4), 931-954.

Athey, S., Tibshirani, J., & Wager, S. (2019). Generalized random forests. *The Annals of Statistics*, 47(2), 1148-1178.

# Function Reference Summary

## Core Estimation Functions

| Function | Purpose |
|----------|---------|
| `ensemble_hte()` | Fit ensemble model for heterogeneous treatment effects |
| `ensemble_pred()` | Fit ensemble model for prediction (no treatment) |
| `combine_ensembles()` | Combine multiple ensemble fits |

## Analysis Functions (Treatment Effects)

| Function | Purpose |
|----------|---------|
| `gates()` | Group Average Treatment Effects |
| `blp()` | Best Linear Predictor of CATE |
| `clan()` | Classification Analysis (characterize groups) |
| `gates_restricted()` | Compare unrestricted vs. restricted targeting |

## Analysis Functions (Prediction)

| Function | Purpose |
|----------|---------|
| `gavs()` | Group Averages (for predictions) |
| `blp_pred()` | Best Linear Predictor (for predictions) |
| `gavs_restricted()` | Compare unrestricted vs. restricted prediction |

## Utility Functions

| Function | Purpose |
|----------|---------|
| `ite()` | Extract ITE prediction matrix from an `ensemble_hte` fit |
| `ensemble_news()` | Check for package updates and view changelog |

## Print and Plot Methods

All results objects have `print()` and `plot()` methods for easy interpretation and visualization.
