---
title: "Quick Guide to ensembleHTE"
author: "Bruno Fava"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Quick Guide to ensembleHTE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
PRECOMPUTED <- file.exists("precomputed/fit_hte_quick.rds")
```

## Introduction

The `ensembleHTE` package provides tools for learning features of heterogeneous treatment effects in randomized controlled trials using an ensemble approach. The package implements the estimation strategy developed in Fava (2025), which combines predictions from multiple machine learning algorithms using Best Linear Predictor (BLP) weights and repeated K-fold cross-fitting.

## Key Concepts

### GATES (Group Average Treatment Effects)

GATES provide a way to test for and visualize treatment effect heterogeneity by:

1. Using machine learning to predict individual treatment effects based on covariates
2. Sorting individuals into groups (e.g., terciles) based on these predictions
3. Estimating the average treatment effect within each group
4. Testing whether high-prediction and low-prediction groups have different effects

### The Ensemble Approach

This package improves upon existing approaches by:

1. **Combining Multiple ML Algorithms**: Combines predictions from several algorithms (Random Forest, XGBoost, etc.) using BLP weights
2. **K-fold Cross-Fitting**: Trains models on K-1 folds and predicts on the held-out fold
3. **Full-Sample Inference**: Uses the entire dataset for final estimation
4. **Repeated Sample-Splitting**: Averages across M repetitions for stability

## Basic Workflow

### 1. Load Data

We use the `microcredit` dataset that ships with the package. It contains data
from a randomized microfinance experiment in the Philippines where microloans
were randomly offered to applicants.

```{r load-data}
library(ensembleHTE)
data(microcredit)

dim(microcredit)

# Variables used throughout this guide
hte_covars <- c("css_creditscorefinal", "own_anybus",
                "max_yearsinbusiness", "css_assetvalue")
has_loan <- microcredit$treat == 1 & microcredit$loan_size > 0

# Income quintiles (used for restricted comparisons)
microcredit$hhinc_quintile <- cut(
  microcredit$hhinc_yrly_base,
  breaks = quantile(microcredit$hhinc_yrly_base, probs = seq(0, 1, by = 0.2)),
  labels = paste0("Q", 1:5),
  include.lowest = TRUE
)
```

### 2. Fit Ensemble HTE Model

We test whether the effect of being offered a microloan on business expenses
varies across borrowers.

```{r fit-model, eval = FALSE}
# Select covariates
hte_covars <- c("css_creditscorefinal", "own_anybus",
                "max_yearsinbusiness", "css_assetvalue")

# Fit ensemble HTE model
fit <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),
  M = 10,   # Use M >= 100 for final results
  K = 4
)

# View results
print(fit)
summary(fit)
```

```{r fit-model-output, echo = FALSE, eval = PRECOMPUTED}
fit <- readRDS("precomputed/fit_hte_quick.rds")
print(fit)
summary(fit)
```

### 3. Analyze Treatment Effect Heterogeneity

```{r analysis, eval = PRECOMPUTED, fig.alt = c("GATES coefficient plot", "CLAN coefficient plot")}
# Best Linear Predictor (BLP)
blp_results <- blp(fit)
print(blp_results)

# Group Average Treatment Effects (GATES)
gates_results <- gates(fit, n_groups = 3)
print(gates_results)
plot(gates_results)

# Classification Analysis (CLAN)
clan_results <- clan(fit, n_groups = 3)
print(clan_results)
plot(clan_results)
```

## Prediction Tasks

The package also supports standard prediction tasks (without treatment effects).
Here we predict bank profitability per peso lent, training only on borrowers
who actually took a loan (where bank profits are observed):

```{r prediction, eval = FALSE}
# Identify borrowers with observed bank profits
has_loan <- microcredit$treat == 1 & microcredit$loan_size > 0

# Fit ensemble prediction model
pred_fit <- ensemble_pred(
  Y = "bank_profits_pp",
  X = c("css_creditscorefinal", "own_anybus", "css_assetvalue",
        "age", "gender", "hhinc_yrly_base"),
  data = microcredit,
  train_idx = has_loan,
  algorithms = c("lm", "grf"),
  M = 10, K = 4
)

# Analyze predictions
summary(pred_fit)

# Group Averages
gavs_results <- gavs(pred_fit, n_groups = 3)
print(gavs_results)
plot(gavs_results)
```

```{r prediction-output, echo = FALSE, eval = PRECOMPUTED, fig.alt = "GAVS bar chart of group averages"}
pred_fit <- readRDS("precomputed/fit_pred_quick.rds")
summary(pred_fit)
gavs_results <- gavs(pred_fit, n_groups = 3)
print(gavs_results)
plot(gavs_results)
```

## Comparing Ranking Strategies

Use `gates_restricted()` or `gavs_restricted()` to compare unrestricted vs.
restricted ranking. Here we test whether requiring income-balanced loan
allocation reduces the lender's ability to identify profitable borrowers:

```{r compare, eval = PRECOMPUTED, fig.alt = "Comparison of unrestricted vs restricted ranking strategies"}
# Compare targeting strategies
comparison <- gavs_restricted(
  pred_fit,
  restrict_by = "hhinc_quintile",
  n_groups = 3,
  subset = "train"
)
print(comparison)
plot(comparison)
```

## Using Subsets

All analysis functions (`gates`, `blp`, `clan`, `gavs`, and their comparison versions) support a `subset` argument that allows you to evaluate results on a subset of observations. This is useful when:

- The outcome is only observed for certain observations
- You want to focus on a specific subpopulation
- You need to exclude certain observations from analysis

```{r subset-examples, eval = PRECOMPUTED}
# Subset with logical vector (e.g., only business owners)
subset_biz <- microcredit$own_anybus == 1
gavs_results <- gavs(pred_fit, n_groups = 3, subset = subset_biz)

# Subset with integer indices
subset_indices <- which(microcredit$lower_window == 1)
gates_results <- gates(fit, n_groups = 3, subset = subset_indices)

# Explicitly use all observations
blp_results <- blp(fit, subset = "all")
```

**Important notes:**
- For `gates()` and `blp()`, the subset must include observations from both treatment and control groups
- For `gavs()`, this is useful when outcomes are only observed for certain units (e.g., treatment effects on outcomes only observable under treatment)
- When using `subset`, the `group_on` argument controls which observations define the group cutoffs. The default (`group_on = "auto"`) uses the same population that was used to **train** the ML model: all observations for `ensemble_hte` fits, or the `train_idx` subset for `ensemble_pred` fits trained on a subset. Set `group_on = "all"` to always use all observations, or `group_on = "analysis"` to form groups within the analysis subset. The same applies to `gates_restricted()` and `gavs_restricted()`.
- A message will be printed indicating how many observations are used when a subset is applied

## Tips and Best Practices

1. **Number of algorithms**: Start with 2-4 diverse algorithms
2. **Number of folds**: K=3 or K=4 recommended
3. **Number of groups**: J=3 (terciles) is standard
4. **Repetitions**: M=5-10 for exploration, M>=100 for final results
5. **Sample size**: Works best with n > 200

## Panel Data

For panel or longitudinal data (multiple observations per individual), use the
`individual_id` argument. This ensures proper fold splitting (all observations
for an individual stay in the same fold) and uses cluster-robust standard errors.
The `microcredit` dataset is cross-sectional, so this example uses simulated data:

```{r panel-data, eval = FALSE}
# Simulate panel data: 200 individuals x 4 periods
set.seed(123)
N <- 200; T_periods <- 4
panel_data <- data.frame(
  id     = rep(1:N, each = T_periods),
  period = rep(1:T_periods, times = N),
  D      = rep(rbinom(N, 1, 0.5), each = T_periods),
  X1     = rnorm(N * T_periods),
  X2     = rnorm(N * T_periods)
)
panel_data$Y <- 2 + panel_data$D * (1 + panel_data$X1) +
  0.5 * panel_data$X2 + rnorm(N * T_periods)

# Fit with individual_id for cluster-aware cross-fitting
fit <- ensemble_hte(
  Y = "Y",
  D = "D",
  X = c("X1", "X2"),
  data = panel_data,
  individual_id = "id",
  algorithms = c("lm", "grf"),
  M = 10, K = 3
)

# All downstream analyses automatically use clustered SEs
gates(fit, n_groups = 3)
blp(fit)
clan(fit)
```

See the complete guide vignette for a full panel data example.

## Performance Tips

Computation time scales with `M * K * length(algorithms)`. Some guidelines:

- **Debugging**: Use `M = 2`, `K = 3`, and 1-2 algorithms to verify your code quickly
- **Exploration**: Use `M = 5-10` with your full algorithm set
- **Final results**: Use `M >= 100` for valid inference (this is required by the methodology)
- **Parallelization**: Set `n_cores > 1` to run repetitions in parallel
- **Tuning**: `tune = TRUE` significantly increases time; use only for final results


## References
Fava, B. (2025). Training and Testing with Multiple Splits: A Central Limit Theorem for Split-Sample Estimators. *arXiv preprint arXiv:2511.04957*.

## Getting Help

- See `?ensemble_hte` and `?gates` for detailed function documentation
- Report issues at: https://github.com/bfava/ensembleHTE/issues
