---
title: "Introduction to ensembleHTE: Detecting Heterogeneity and Exploring Targeting Tradeoffs"
author: "Bruno Fava"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
# Pre-computed fit objects are stored in vignettes/precomputed/.
# Run data-raw/precompute_walkthrough.R to regenerate them.
PRECOMPUTED <- file.exists("../precomputed/fit_hte_walkthrough.rds")
```

## Introduction

The `ensembleHTE` package provides a practical approach to heterogeneous
treatment effect (HTE) analysis and prediction that (i) uses the entire
sample for both training and evaluation through repeated cross-fitting, and
(ii) combines predictions from multiple machine learning algorithms into a
single ensemble. It handles both treatment-effect estimation
(`ensemble_hte`) and standard prediction tasks (`ensemble_pred`).
A suite of downstream analysis functions—BLP, GATES, CLAN and GAVS,
originally proposed by
[Chernozhukov et al. (2025)](https://www.econometricsociety.org/publications/econometrica/2025/07/01/FisherSchultz-Lecture-Generic-Machine-Learning-Inference-on-Heterogeneous-Treatment-Effects-in-Randomized-Experiments-With-an-Application-to-Immunization-in-India) and adapted
here following
[Fava (2025)](https://arxiv.org/abs/2511.04957)—lets you test for
heterogeneity, characterize who is in each group, and compare
targeting strategies.

This walkthrough uses data from a randomized microfinance experiment in the
Philippines, first studied by
[Karlan and Zinman (2011)](https://doi.org/10.1126/science.1200138),
in which individual-liability microloans were randomly offered to applicants.
The analysis is inspired by
[Athey, Fava, Karlan, Osman and Zinman (2025)](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf),
which studies whether targeting microloans for bank profitability comes at
the cost of borrower welfare. We use the dataset as a running example to
illustrate the package's main features through three exercises:

1. **Do microloans affect all borrowers equally?** We use `ensemble_hte()` to
   detect and characterize heterogeneous treatment effects on business
   expenses.
2. **Is there a tradeoff between bank profits and borrower welfare?** We use
   `ensemble_pred()` to predict which borrowers are most profitable for the
   bank, then ask whether those same borrowers also benefit the most—or
   whether lending to profitable borrowers comes at borrowers' expense.
3. **What if the lender must maintain income balance?** Targeting loans based
   on predicted profitability tends to steer credit toward higher-income
   applicants. We use the restricted analysis functions to ask whether
   requiring balance across household-income quintiles meaningfully reduces
   the lender's ability to identify profitable borrowers.

> **Note on `M`:** Throughout this walkthrough we use `M = 10` repetitions
> for speed. For actual applications, I strongly recommend using `M >= 100`
> to ensure stable inference. You can start with a small `M` to verify your
> code, then scale up. If computation takes a long time,
> `combine_ensembles()` lets you split the work across sessions and merge
> the results afterward.

## The data

The `microcredit` dataset ships with the package. Treatment assignment
depended on applicants' credit scores: near a scoring cutoff, applicants
were randomly assigned to receive a loan offer or not. A follow-up survey
measured outcomes 11–22 months later.

```{r load-data}
library(ensembleHTE)
data(microcredit)

dim(microcredit)
```

Key variables for our analysis:

| Variable                  | Description |
|---------------------------|-------------|
| `treat`                   | Treatment indicator (1 = loan offered) |
| `prop_score`              | Known propensity score (varies by credit-score window) |
| `exp_yrly_end`            | Yearly business expenses at endline (pesos) |
| `hhinc_yrly_end`          | Yearly household income at endline (pesos) |
| `bank_profits_pp`         | Bank profit per peso lent (only meaningful for borrowers) |
| `loan_size`               | Loan amount (0 if no loan taken) |
| `css_creditscorefinal`    | Credit score at application |
| `lower_window`            | Indicator for the lower credit-score window |
| `own_anybus`              | Owns any business (1 = yes) |
| `max_yearsinbusiness`     | Maximum years in business |
| `css_assetvalue`          | Business asset value at application |

Because treatment was randomized within credit-score windows, the propensity
score varies across individuals and is provided in the data. We pass it
explicitly to every function that needs it.

```{r data-overview}
# Treatment shares
table(microcredit$treat)

# Propensity score summary
summary(microcredit$prop_score)
```

---

## Part 1 — Do microloans affect all borrowers equally?

### Setting up the specification

We want to test whether the effect of being offered a microloan on **business
expenses** varies across borrowers. We select a small set of covariates that
capture the applicant's financial profile and business background. There is
nothing special about this particular selection—it is chosen for convenience
and ease of exposition. The original paper uses a broader set of covariates
(available as the `microcredit_covariates` object in the package).

```{r hte-covars}
hte_covars <- c(
  "css_creditscorefinal",   # credit score
  "lower_window",           # lower credit-score window indicator
  "own_anybus",             # owns any business
  "max_yearsinbusiness",    # years of business experience
  "css_assetvalue"          # business asset value
)
```

### Fitting the ensemble

`ensemble_hte()` trains an ensemble of ML algorithms to detect heterogeneous
treatment effects. It uses repeated cross-fitting—controlled by `M`
(number of repetitions) and `K` (number of folds)—to produce stable
predictions and valid standard errors.

The example below uses `M = 10` repetitions and two algorithms: OLS (`lm`)
and generalized random forests (`grf`). We discuss a preferred
specification at the end of this section.

```{r hte-fit, eval = FALSE}
set.seed(2026)

fit_hte <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("lm", "grf"),
  M = 10,
  K = 4
)
```

```{r hte-fit-load, echo = FALSE, eval = PRECOMPUTED}
fit_hte <- readRDS("../precomputed/fit_hte_walkthrough.rds")
```

The `print()` method confirms the specification:

```{r hte-print, eval = PRECOMPUTED}
print(fit_hte)
```

And `summary()` gives a first overview of the results:

```{r hte-summary, eval = PRECOMPUTED}
summary(fit_hte)
```

The summary reports two key diagnostics. `beta1` (the average treatment
effect) is −2,322 pesos with a p-value of 0.59—no evidence of an overall
effect of loan offers on business expenses. `beta2` (the heterogeneity
loading) is 0.71 with a p-value of 0.02—significant evidence that the ML
model is well calibrated for heterogeneity: predicted treatment effects are
correlated with actual treatment effects. The GATES top-minus-bottom gap is
about 24,800 pesos (p = 0.03), reinforcing this finding.

Let's look at each of these in more detail.

### Is there heterogeneity? BLP

The **Best Linear Predictor (BLP)** asks two questions:

- **Is there an average treatment effect?** `beta1` tests whether the loan
  offer shifts business expenses on average.
- **Does the ML model capture real heterogeneity?** `beta2` tests whether the
  predicted treatment effects are correlated with actual treatment effects—a
  significant result means people are genuinely affected differently.

```{r blp, eval = PRECOMPUTED}
blp_res <- blp(fit_hte)
print(blp_res)
```

```{r blp-plot, eval = PRECOMPUTED, fig.cap = "BLP: average treatment effect and heterogeneity loading."}
plot(blp_res)
```

Here, `beta1` is not significant (p = 0.59): there is no detectable average
effect of loan offers on business expenses. `beta2` is significant
(p = 0.02): the ML predictions are correlated with actual heterogeneity,
meaning that the effect of the loan offer genuinely differs across
borrowers.

### How do effects vary? GATES

**GATES (Group Average Treatment Effects)** sorts individuals into groups
based on their predicted treatment effect—from lowest to highest—and
estimates the average effect in each group. If the treatment truly affects
people differently, these group-level effects should vary systematically.

```{r gates, eval = PRECOMPUTED}
gates_res <- gates(fit_hte, n_groups = 3)
print(gates_res)
```

```{r gates-plot, eval = PRECOMPUTED, fig.cap = "GATES: average treatment effect by predicted-effect tercile, with 95% confidence intervals."}
plot(gates_res)
```

The bottom tercile—the third of borrowers predicted to have the lowest
treatment effects—shows an average treatment effect of −15,540 pesos,
meaning the loan offer is estimated to *reduce* their business expenses.
The middle tercile is near zero (−679), and the top tercile shows a
positive effect of +9,226 pesos. None of these is individually significant
at the 5% level, which is not surprising given the sample size. But the
pattern is strongly monotonic, and the top-minus-bottom difference of
24,766 pesos—saying the average treatment effect is 24,800 pesos higher in
the top group than in the bottom group—is statistically significant
(p = 0.03). This is consistent with the BLP result: the treatment
genuinely affects borrowers differently.

### Who is in each group? CLAN

Now that we know treatment effects vary, the natural question is *who* is
in each group. **CLAN (Classification Analysis)** answers this by computing
the average of each covariate within the GATES groups and testing whether
those averages differ between the top and bottom groups.

```{r clan, eval = PRECOMPUTED}
clan_res <- clan(fit_hte, n_groups = 3)
print(clan_res)
```

```{r clan-plot, eval = PRECOMPUTED, fig.cap = "CLAN: covariate means by GATES group (scaled for comparability)."}
plot(clan_res)
```

Compared to the bottom group (those whose business expenses are predicted to
decrease most), borrowers in the top group:

- Have significantly **higher business asset values** (93,546 vs. 53,070
  pesos, p < 0.01).
- Have significantly **fewer years in business** (5.3 vs. 8.6 years,
  p < 0.001).
- Are significantly **less likely to already own a business** (62% vs. 84%,
  p < 0.001).
- Have slightly **lower credit scores** (50.5 vs. 52.1, p < 0.001).

Borrowers who increase their expenditures most in response to the loan
offer tend to be asset-rich but relatively new to business
ownership—perhaps people with resources who are using the microloan to
start or expand a business. By contrast, those in the bottom group are
longer-established business owners with higher credit scores.

### Preferred specification

The example above uses `M = 10` and two algorithms for speed. For a more
thorough analysis, I recommend using `M >= 100` repetitions (even larger
values are better if computationally feasible—
[Athey et al. (2025)](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf)
uses `M = 500`), and a richer set of algorithms: generalized random forests
(`grf`), elastic net (`glmnet`), neural networks (`nnet`), and gradient
boosting (`xgboost`). More repetitions improve the stability of the
inference, and more algorithms give the ensemble a wider range of modeling
strategies to draw from.

```{r hte-fit-recommended, eval = FALSE}
fit_hte <- ensemble_hte(
  Y = "exp_yrly_end",
  D = "treat",
  X = hte_covars,
  data       = microcredit,
  prop_score = "prop_score",
  algorithms = c("grf", "glmnet", "nnet", "xgboost"),
  M = 200,
  K = 4
)
```

If computation takes a long time, `combine_ensembles()` lets you split the
work across sessions and merge the results afterward.

---

## Part 2 — Is there a tradeoff between bank profits and borrower welfare?

### The question

Beyond impacts on borrowers, lenders and policymakers care about the financial
sustainability of microlending. Can we identify which borrowers are most
profitable for the bank? And if the lender targets the most profitable
borrowers, does that come at borrowers' expense—or can both goals coexist?

This is a central question in
[Athey, Fava, Karlan, Osman and Zinman (2025)](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf).
Notice that this is not a treatment-effect question: we are simply predicting
an outcome (bank profits), then examining what that prediction implies for
other outcomes. The `ensemble_pred()` function handles this type of
prediction task.

### Predicting an outcome that is only partially observed

Bank profit per peso lent (`bank_profits_pp`) is only observed for individuals
who were offered a loan *and* actually took it—the bank only earns revenue on
disbursed loans. For everyone else, bank profits are undefined.

`ensemble_pred()` handles this through the `train_idx` argument: we tell the
function to **train** on the subset where profits are observed, while still
generating **predictions for everyone**. This matters because we want to rank
*all* applicants by predicted profitability, not just those who happened to
get a loan.

```{r pred-setup}
# Who has observable bank profits?
has_loan <- microcredit$treat == 1 & microcredit$loan_size > 0
sum(has_loan)
```

There are 650 borrowers with observed bank profits. The model will train on
these but predict profitability for all 1,113 applicants.

### Choosing covariates

We pick covariates plausibly related to bank profitability. As before, this is
a small selection for illustration; the original paper uses a richer
specification.

```{r pred-covars}
profit_covars <- c(
  "css_creditscorefinal",   # credit score
  "lower_window",           # credit-score window
  "own_anybus",             # business ownership
  "max_yearsinbusiness",    # business experience
  "css_assetvalue",         # business asset value
  "css_stockvalue",         # stock value
  "age",                    # age
  "gender",                 # gender
  "hhinc_yrly_base",        # baseline household income
  "exp_yrly_base"           # baseline business expenses
)
```

### Fitting the ensemble

```{r pred-fit, eval = FALSE}
set.seed(2026)

fit_pred <- ensemble_pred(
  Y    = "bank_profits_pp",
  X    = profit_covars,
  data = microcredit,
  train_idx  = has_loan,
  algorithms = c("lm", "grf"),
  M = 10,
  K = 4
)
```

```{r pred-fit-load, echo = FALSE, eval = PRECOMPUTED}
fit_pred <- readRDS("../precomputed/fit_pred_walkthrough.rds")
```

As with `ensemble_hte()`, using `M >= 100` repetitions and a richer set of
algorithms (`grf`, `glmnet`, `nnet`, `xgboost`) is recommended for real
applications.

```{r pred-print, eval = PRECOMPUTED}
print(fit_pred)
```

```{r pred-summary, eval = PRECOMPUTED}
summary(fit_pred)
```

The summary tells us the model has meaningful predictive power: the R-squared
on the training observations is 0.22 and the correlation between predictions
and outcomes is 0.47. The built-in BLP diagnostic shows an intercept close to
zero (−0.01, p = 0.99) and a slope close to one (0.99, p < 0.001), meaning
the predictions are well calibrated.

The GAVS section at the bottom already shows the model successfully separates
borrowers: the bottom tercile averages −10.0 pesos of profit per peso lent,
while the top tercile averages +8.7. The top-minus-bottom gap of 18.7 pesos
is highly significant (p < 0.001).

Let's look at each of these in more detail.

### Does the model predict well? BLP for predictions

Let's examine the calibration diagnostic more closely. `blp_pred()` is the
prediction analogue of `blp()`. It regresses the observed outcome on the ML
predictions and tests two things:

- **Is the intercept close to zero?** If so, the predictions are not
  systematically biased.
- **Is the slope close to one?** If so, the predictions are well calibrated:
  a one-unit increase in predicted profit corresponds to a one-unit increase
  in actual profit.

```{r blp-pred, eval = PRECOMPUTED}
blp_pred_res <- blp_pred(fit_pred)
print(blp_pred_res)
```

The intercept is −0.01 (not significantly different from zero) and the slope
is 0.99 (highly significant and close to one). This confirms the model is
well calibrated.

### Group averages of bank profits: GAVS

**GAVS (Group Averages)** sorts observations into groups by their predicted
bank profit and computes the average *observed* profit in each group. If the
model is useful, higher-prediction groups should have higher observed profits.

Because `bank_profits_pp` is only observed for borrowers who took a loan,
`gavs()` automatically restricts the analysis to those observations:

```{r gavs, eval = PRECOMPUTED}
gavs_res <- gavs(fit_pred, n_groups = 3)
print(gavs_res)
```

```{r gavs-plot, eval = PRECOMPUTED, fig.cap = "GAVS: average observed bank profits by predicted-profit tercile."}
plot(gavs_res)
```

The results confirm the model sharply separates profitable from unprofitable
borrowers. The bottom tercile (those predicted to be least profitable) has an
average observed profit of −10.0 pesos per peso lent—the bank loses money on
these borrowers. The middle tercile is roughly break-even (−1.5, not
significant). The top tercile earns +8.7 pesos per peso lent. The
top-minus-bottom gap of 18.7 pesos is highly significant (p < 0.001).

### Do profitable borrowers also benefit from the loan? GATES

This is the key question. We use `gates()` with a twist: we keep the groups
defined by *predicted bank profits*, but we switch the outcome to *household
income*. In other words, among borrowers the bank would rank as most
profitable, what happens to their income when they are offered a loan?

Because household income is observed for everyone (treated and control), no
subsetting is needed here. We pass the `outcome`, `treatment`, and
`prop_score` arguments to tell `gates()` to estimate treatment effects on a
different outcome than the one the model was trained on:

```{r gates-pred, eval = PRECOMPUTED}
gates_hhinc <- gates(
  fit_pred,
  outcome    = "hhinc_yrly_end",
  treatment  = "treat",
  prop_score = "prop_score",
  n_groups   = 3
)
print(gates_hhinc)
```

```{r gates-pred-plot, eval = PRECOMPUTED, fig.cap = "GATES: treatment effect on household income, by predicted bank-profit tercile."}
plot(gates_hhinc)
```

How to read this:

- If the top group (most profitable for the bank) also shows the **largest
  positive** treatment effect on income, profitability and welfare are
  **complements**—the lender can do well and do good at the same time.
- If the top group shows a **smaller or negative** effect, there is a
  **tradeoff**—profitable borrowers are not the ones who benefit most.
- If there is no significant difference across groups, profitability and
  welfare may be unrelated—but this could also reflect low statistical
  power rather than a true absence of a relationship.

The point estimates suggest a possible complementarity: the top profit group
shows a treatment effect on household income of +5,471 pesos, while the
bottom and middle groups are near zero (−704 and −453, respectively). However,
none of these estimates is individually significant, and the top-minus-bottom
gap of 6,175 pesos has a p-value of 0.21. With this sample size and `M = 10`,
we do not have enough power to draw a firm conclusion about whether
profitability and welfare are complements or substitutes. The lack of a
significant relationship could genuinely mean no tradeoff exists, but it could
also simply reflect insufficient statistical power—too much noise in the
treatment-effect estimates given the available sample.

### Who is in each profit group? CLAN

```{r clan-pred, eval = PRECOMPUTED}
clan_pred <- clan(fit_pred, n_groups = 3, variables = profit_covars)
print(clan_pred)
```

```{r clan-pred-plot, eval = PRECOMPUTED, fig.cap = "CLAN: covariate means by predicted-profit group (scaled for comparability)."}
plot(clan_pred)
```

The CLAN results reveal a crucial tradeoff between lender profits and
distributional objectives—a central finding of
[Athey et al. (2025)](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf).
Compared to the bottom tercile, the top profit group:

- Has **higher credit scores** (52.5 vs. 48.7, p < 0.001) and is less likely
  to be in the lower credit-score window (9% vs. 27%).
- Has **much higher business assets** (112,626 vs. 19,870 pesos) and **stock
  values** (107,578 vs. 12,567 pesos)—both p < 0.001.
- Has **higher baseline income** (30,404 vs. 9,972 pesos) and **business
  expenses** (128,420 vs. 40,939 pesos)—both p < 0.001.
- Is **younger** (40.3 vs. 44.1 years, p < 0.001) and more likely to be
  **male** (80% vs. 91% female, p < 0.001).
- Is **less likely to own a business** at baseline (59% vs. 84%, p < 0.001).

Even though the GATES analysis above found no evidence of a tradeoff between
profits and *treatment effects* on income, the CLAN results reveal an
important distributional tradeoff: selecting the most profitable borrowers
effectively means selecting wealthier, higher-income, better-scored
applicants with a higher proportion of men. This motivates Part 3: if the
bank targets profitability, what happens when a regulator requires that loan
allocation remain balanced across income groups?

---

## Part 3 — What if the lender must maintain income balance?

### The problem

The CLAN results from Part 2 showed that profitable borrowers tend to have
substantially higher baseline incomes (30,404 vs. 9,972 pesos). If a lender
targets loans based on predicted profitability, it will systematically direct
credit toward higher-income applicants—the very people who are already better
off. A regulator or the lender itself might respond with a constraint: loan
allocation must be balanced across household-income groups.

But does the equity constraint meaningfully reduce the lender's ability to
identify profitable borrowers—or can both goals coexist?

### How restricted targeting works

The `gavs_restricted()` and `gates_restricted()` functions compare two
targeting strategies:

- **Unrestricted**: rank all applicants by predicted profitability and form
  groups from the overall distribution.
- **Restricted**: rank applicants *within* each income quintile, then form
  groups within each stratum. This ensures every income group contributes
  proportionally to the top, middle, and bottom profit groups.

The difference between the two measures the efficiency cost of the constraint.

### Creating income quintiles

We split borrowers into five equally sized groups (quintiles) based on their
baseline household income. The `cut()` function assigns each borrower to a
quintile by computing the 20th, 40th, 60th, and 80th percentiles of
`hhinc_yrly_base` and labeling each observation Q1 (lowest income) through
Q5 (highest income):

```{r restrict-setup, eval = PRECOMPUTED}
hhinc_quintile <- cut(
  microcredit$hhinc_yrly_base,
  breaks  = quantile(microcredit$hhinc_yrly_base, probs = seq(0, 1, by = 0.2)),
  labels  = paste0("Q", 1:5),
  include.lowest = TRUE
)

table(hhinc_quintile)
```

### Does the constraint hurt profitability? GAVS restricted

We compare average observed bank profits across terciles under both strategies:

```{r gavs-restricted, eval = PRECOMPUTED}
gavs_r <- gavs_restricted(
  fit_pred,
  restrict_by = hhinc_quintile,
  n_groups    = 3,
  subset      = "train"
)
print(gavs_r)
```

```{r gavs-restricted-plot, eval = PRECOMPUTED, fig.cap = "Unrestricted vs. restricted GAVS on bank profits."}
plot(gavs_r)
```

The results show that the equity constraint comes at a real cost to
profitability targeting. Under unrestricted targeting, the top-minus-bottom
gap in observed bank profits is 18.7 pesos per peso lent. Under the
income-balanced constraint, this gap shrinks to 9.0—roughly half. The
difference of 9.7 pesos is highly significant (p < 0.001).

Notice that the restricted model can still identify profitable borrowers:
the top tercile under the equity constraint averages +4.9 pesos per peso
lent, and the bottom tercile −4.1. The constraint does not eliminate the
model's predictive power—it just reduces it.

This makes intuitive sense: if the model uses income-related covariates to
predict profitability, requiring balance across income quintiles removes a
key source of predictive variation. The lender can still identify profitable
borrowers within each income group, but the overall separation is
substantially reduced.

### Does the constraint change welfare impacts? GATES restricted

The same comparison applies to treatment effects on household income. Do the
welfare consequences of profit-based targeting change under the equity
constraint?

```{r gates-restricted, eval = PRECOMPUTED}
gates_r <- gates_restricted(
  fit_pred,
  restrict_by = hhinc_quintile,
  outcome     = "hhinc_yrly_end",
  treatment   = "treat",
  prop_score  = "prop_score",
  n_groups    = 3
)
print(gates_r)
```

```{r gates-restricted-plot, eval = PRECOMPUTED, fig.cap = "Unrestricted vs. restricted GATES on household income."}
plot(gates_r)
```

Under unrestricted targeting, the top profit group shows a treatment effect on
household income of +5,471 pesos (the complementarity pattern from Part 2).
Under restricted targeting, this drops to +2,254, and the top-minus-bottom gap
shrinks from 6,175 to 2,670 pesos. However, none of the differences between
unrestricted and restricted targeting are statistically significant (p = 0.31
for the top-minus-bottom gap). This is largely because the treatment effect
estimates themselves are noisy—as we saw in Part 2, individual group effects
are not significant even under unrestricted targeting.

In practical terms: the income-balance constraint clearly reduces the lender's
ability to sort borrowers by profitability, but we cannot say with confidence
whether it changes the welfare implications of that sorting.

> **Note:** `restrict_by` accepts either a character scalar naming a column
> in the data (e.g., `restrict_by = "hhinc_quintile"`) or a vector of the
> same length as the data. For continuous variables, use `cut()` to create
> bins first, as we did above with income quintiles.

---

## Summary

| Exercise | Question | Key functions |
|----------|----------|---------------|
| **Part 1** | Do microloans affect all borrowers equally? | `ensemble_hte()` → `blp()`, `gates()`, `clan()` |
| **Part 2** | Is there a tradeoff between bank profits and borrower welfare? | `ensemble_pred()` → `blp_pred()`, `gavs()`, `gates()`, `clan()` |
| **Part 3** | What happens if the lender must maintain income balance? | `gavs_restricted()`, `gates_restricted()` |

A few things worth remembering:

- **`train_idx`** is essential when your outcome is only observed for a
  subset of individuals. The models train on that subset but predict for
  everyone.
- **`gates()` can use a different outcome** than the one the model was
  trained on. This is how we asked: "among the groups formed by predicted bank
  profits, what are the treatment effects on household income?" Just pass
  the `outcome`, `treatment`, and `prop_score` arguments.
- **`ensemble_hte()` objects can also use `gavs()`** with a different outcome.
  For example, after estimating heterogeneous treatment effects on business
  expenses, you could use `gavs(fit_hte, outcome = "bank_profits_pp")` to
  ask whether borrowers predicted to increase their expenditures more also
  generate higher profits for the bank—another way to explore potential
  tradeoffs.
- **`restrict_by`** lets you compare free targeting against
  equity-constrained targeting. Pass any categorical variable to define the
  strata.
- **Use `M >= 100`** for publication-quality inference. The `M = 10` used
  here is for illustration only. Larger values are even better if
  computationally feasible—[Athey et al. (2025)](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf)
  uses `M = 500`. `combine_ensembles()` can merge results across sessions if
  computation time is a concern.

## References

- Athey, S., Fava, B., Karlan, D., Osman, A. and Zinman, J. (2025). *Profits
  and Social Impacts: Complements vs. Tradeoffs for Lenders in Three
  Countries*. Working paper.
  [Link](https://bfava.com/files/Tradeoffs_in_Microcredit.pdf)

- Chernozhukov, V., Demirer, M., Duflo, E. and Fernandez-Val, I. (2025).
  Fisher–Schultz Lecture: Generic Machine Learning Inference on Heterogeneous Treatment Effects in Randomized Experiments, With an Application to Immunization in India.
  *Econometrica*, **93**(4), 1121–1164.
  [doi:10.3982/ECTA19303](https://doi.org/10.3982/ECTA19303)

- Fava, B. (2025). Training and Testing with Multiple Splits: A Central Limit
  Theorem for Split-Sample Estimators. *arXiv preprint arXiv:2511.04957*.
  [Link](https://arxiv.org/abs/2511.04957)

- Karlan, D. and Zinman, J. (2011). Microcredit in Theory and Practice: Using
  Randomized Credit Scoring for Impact Evaluation. *Science*, **332**(6035),
  1278–1284. [doi:10.1126/science.1200138](https://doi.org/10.1126/science.1200138)
